---
title: "CosMx Data Insitutype Unsupervised Clustering and scType annotation"
author: 
  - name: "Lidia Getino Álvarez"
    affiliation: "M0.207 TFM - MU Bioinformática y Bioestadística (UOC)"
date: "`r format(Sys.Date(), '%d %B %Y')`"
site: workflowr::wflow_site
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

<!-- This pipeline has been developed using Seurat v4.4, as Seurat v5 presents several changes to the object structure that affected the working methods. Aditionally, the AtoMx exported Seurat objects come in the Seurat v4 structure. Therefore, working in this version can be more compatible with exported data too. -->

```{r setup, include = FALSE}
# Read utils
source(
  here::here("code", "utils.R"), 
  local = knitr::knit_global())

source(
  here::here("code", "aux_functions.R"), 
  local = knitr::knit_global())

# Setup chunk options
## By reading *utils.R*, the *chunk hook* function "monitor" can be included
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  monitor = TRUE, # Self created chunk hook to monitor time and memory usage
  error = TRUE
)
```

# Dependencies

```{r Libraries, message = FALSE}
library(data.table) # Efficient data management
library(Matrix) # Sparse matrices
library(here) # Enhanced file referencing in project-oriented workflows
library(dplyr) # For the use of pipes %>%
library(kableExtra) # For table formatting
library(Seurat) # Seurat object
library(ggplot2) # Graphics
library(patchwork) # Layout graphics
library(pheatmap) # Heatmap plot
library(HGNChelper) # For scType annotation

#remotes::install_github("Nanostring-Biostats/InSituType")
library(InSituType)
```

# Load the data

First of all, data needs to be loaded into the session. For this script, two inputs will be needed:  

* The Seurat object from the "Dimensional reduction" script.  
* The Negative probe matrix from the "Data loading" script.  

<!-- Insitutype algorithm uses not normalized raw counts to cluster the cells. However, the UMAP reduction will be used for plot visualizations, and is therefore needed. For the main pipeline I will use the SCT Seurat object. -->

```{r LoadData}
# Indicate the object folder
folder <- "SCT" # Choose between "RC", "Log" or "SCT"

if (!folder %in% c("RC", "Log", "SCT")) {
  stop("The selected folder is invalid, choose: 'RC', 'Log' or 'SCT'")
}

# Load Seurat object
name <- paste0("seu_", folder, "_um.RDS")
seu <- readRDS(here("output","processed_data",folder,name))

# Load negMat
negcounts <- readRDS(here("output","processed_data","negMat_unfiltered.RDS"))

# Load the number of PCs used for dimensional reduction
name <- paste0("npcs_", folder, ".RDS")
npcs <- readRDS(here("output","processed_data",folder,name))
```

In this script, clustering will be done using the unsupervised clustering method by *InSituType*. This method was developed by Nanostring, owner of the CosMx SMI technology, and is, therefore, optimized to work with this type of data specifically.  

# Prepare the data

In order to apply the InSituType algorithm, it is necessary to provide the function with the following information:  

* Raw expression counts matrix (not normalized), with cells in columns and genes in rows.  
* A vector of the mean negative control value per cell.  

```{r PrepData}
# Extract raw counts from the Seurat object
counts <- t(seu@assays[["RNA"]]@counts)

# Note the matrix is transposed to have cells in rows and genes in cols

# Ensure negcounts has the same cells as counts
if (!identical(rownames(counts), rownames(negcounts))) {
    negcounts <- negcounts[rownames(counts),] # Filter and/or re-order if needed
} 

# Calculate Negative control means per cell
negmean <- rowMeans(negcounts)
```

Additionally, the function can take the following optional inputs, if available:  

* A vector of background counts per gene in each cell --> *Insitutype can automatically estimate this with the negative control counts*.   
* A reference profile matrix with genes in rows and cell types in columns --> *This is only necessary for supervised or semi-supervised clustering*.  
* Additional information incorporated as "cohort", such as inmunofluorescence stains or precomputed spatial niches --> *Including this option is low-risk, as it will not be taken into account if it is uninformative*.  

For this example, immunofluoresce data will be included as a cohort.  

<!-- In case no cohort wants to be included, this chunk can be commented so that it is not executed -->

```{r PrepCohort}
# Extract Immunoflurence data (seen in Alex Soupir video's)
imm_fluor <- seu@meta.data %>%
  select(contains(c("Mean","Max"))) %>%
  select(-contains("DAPI"))

# Set as cohort with InSituType function "fastCohorting"
cohort <- fastCohorting(imm_fluor, gaussian_transform = TRUE)

# This function creates "groups" or similar cells based on their immunofluoresce,
# as that is the input I have given it

table(cohort)
```

As it can be observed, the "fastCohorting" function has created "groups" of similar cells based on their immunofluoresce abundances.  

# Unsupervised Insitutype clustering

In this method, the user provides a number or a range of clusters to evaluate. Then the function will cluster the cells on the selected number of clusters.  

```{r Unsup}
## Code adapted from the InSituType vignette

# Unsupervised Insitutye clustering with immunofluoresce cohort
unsup <- insitutype(
  x = counts,
  neg = negmean,
  assay_type = "RNA",
  cohort = cohort,
  bg = NULL, # The function will estimate it automatically
  reference_profiles = NULL, # Not needed for unsupervised clustering
  n_clusts = 5:20, # Will test all these clusters and select the better fit
  n_phase1 = 200,
  n_phase2 = 500,
  n_phase3 = 2000,
  n_starts = 1,
  max_iters = 5
)
```

As a result the function provides a list with:  

* Vector of cluster assignments.  
* Vector of probabilities, or confidence, of each cell's cluster.  
* Matrix of log-likelihoods with cells on rows and clusters in columns.  
* Matrix of cluster-specific expression profiles.  

```{r UnsupProb}
## Code adapted from the InSituType vignette

# Clusters
n_clus <- length(unique(unsup$clust))

# Example of probabilities
round(head(unsup$prob), 2)
summary(unsup$prob)
```
As it can be observed, the algorithm has selected `r n_clus` clusters as the better fit, with the majority of cells presenting a high confidence of belonging to the selected cluster.  

Now, it might be interesting to explore the confidence ratios in order to select an appropriate threshold for low probability assignments, and/or see if the low probabilities are concentrated in any particular cluster.  

## Exploration of cell assignment scores

As per the [InSituType vignette](https://github.com/Nanostring-Biostats/InSituType/blob/main/vignettes/NSCLC-clustering-vignette.Rmd), clusters with an average probability of less than 75% are unreliable. Therefore, for the exploration a threshold of 75% will be selected.  

```{r ExporingUnsupProb}
# Determine initial low confidence cells
thresh <- 0.75 
low_conf_cells <- names(unsup$prob[unsup$prob < thresh])
n_low <- length(low_conf_cells)
perc_low <- round(n_low / length(unsup$prob) * 100, 1)

# Create dataframe for plotting
df <- data.frame(clust = unsup$clust,
                 prob = unsup$prob)
```

```{r VizUnsupProb, fig.width = 10, fig.height = 5}
# Histogram of confidence
p1 <- ggplot(df, aes(x = prob)) +
  geom_histogram(bins = 50, fill = "lightblue", colour = "black") +
  geom_vline(xintercept = thresh, colour = "red", lwd = 1) +
  ggtitle(paste("Nº cells with prob <", thresh, ":", n_low)) +
  labs(title = paste("Nº cells with prob <", thresh, ":", n_low),
       x = "Assignment Confidence (%)",
       y = "Nº Cells")

# Boxplots of confidence by cluster
p2 <- ggplot(df, aes(x = clust, y = prob)) +
  geom_boxplot() +
  geom_hline(yintercept = thresh, colour = "red", lwd = 1) +
  labs(title = "Confidence per Cluster",
       x = "Cluster", 
       y = "Assignment Confidence (%)")

# Arrange plots
p1 + p2
```

All clusters have an average probability of more than 75%, although many outliers are observed. Finally, there are `r n_low` (`r perc_low` %) cells with an assignment confidence of less than `r thresh*100` %, showing that most of the cells have been clustered very confidently.  

## Flightpath exploration 

The InSituType package includes an interesting visualization tool called "flightpath_plot", which represents cells according to their cluster probabilities. This will allow refining the clustering assignation.  

```{r VizFPUnsup, fig.width = 7, fig.height = 7}
## Code adapted from the InSituType vignette

# Create custom color palette
cols <- gg_color_hue(n_clus)
names(cols) <- unique(unsup$clust)

# Flightpath plot
flightpath_plot(insitutype_result = unsup,
                col = cols[unsup$clust])
```

As before, it can be seen that all clusters are very well defined, with average probabilities of nearly 100%. A few connections are observed between some clusters, representing cells whose expression profiles could be confused between said clusters.   

## Clustering refining

After exploration, using the function "refineClusters", the user can determine whether some clusters should be deleted, merged or subclustered.  

However, in this case, all clusters seem very well defined and present good quality assignments scores, so no refining is needed.

```{r RefinedUnsup, include = FALSE}
## Code adapted from the InSituType vignette

# Clustering refining (left as an example, if used, delete "include = FALSE")
# newclusts <- refineClusters(logliks = sup$logliks,
#                             assay_type = "RNA",
#                             merges = NULL,
#                             to_delete = NULL,
#                             subcluster = NULL,
#                             counts = counts,
#                             neg = negmean)
```

## Add clustering to Seurat object

Now, the clusters can be included in the Seurat object for UMAP visualization.  

```{r AddClus}
# Add new metadata column to seurat object
seu$IST_unsup_clus <- as.factor(unsup$clust)
#seu$IST_unsup_clus_refined <- as.factor(newclusts$clust) # Refined clusters

# Set as default idents
Idents(seu) <- seu$IST_unsup_clus
```

```{r VizClus, fig.width = 5, fig.height = 5}
# Create custom color palette
n <- length(unique(Idents(seu)))
cols <- gg_color_hue(n)

# DimPlot of selected resolution
DimPlot(seu, reduction = "umap", label = TRUE, cols = cols, raster = FALSE)
```

Finally, the visualization of "nFeature" and "nCount" allows to identify clusters with better/lower representation by the CosMx panel.  

```{r VizClus2, fig.width = 6, fig.height = 4}
## Code adapted from CosMxLite vignette
## Points are removed as CosMx datasets can contain over 1 million cells and may hide the violin plot

# Violin plot
VlnPlot(seu, features = c("nFeature_RNA", "nCount_RNA"), ncol = 1, pt.size = 0, group.by = "IST_unsup_clus")
```

# Cell annotation - scType

In this script, clusters will be annotated based on their top markers using the "scType" package and its "Brain" reference. For more information about this method check its documentation and vignette  [here](https://github.com/IanevskiAleksandr/sc-type).  

```{r scTypePrep}
## Code adapted from scType vignette

# Load gene set preparation and cell type annotation function
source("https://raw.githubusercontent.com/IanevskiAleksandr/sc-type/master/R/gene_sets_prepare.R")
source("https://raw.githubusercontent.com/IanevskiAleksandr/sc-type/master/R/sctype_score_.R")

# Prepare DB file and select tissue
db_ <- "https://raw.githubusercontent.com/IanevskiAleksandr/sc-type/master/ScTypeDB_full.xlsx";
tissue <- "Brain" # e.g. Immune system, Pancreas, Liver, Kidney, Brain...

# Prepare gene sets
gs_list <- gene_sets_prepare(db_, tissue)

# Extract scaled data from the Seurat object
if (folder == "SCT") {
  slot <- "SCT"
} else {
  slot <- "RNA"  
}

scaled_data <- as.matrix(seu[[paste0(slot)]]@scale.data) 
```

```{r RunscType}
## Code adapted from scType vignette

# Run ScType
es.max <- sctype_score(scaled_data, scaled = TRUE, 
                       gs = gs_list$gs_positive, 
                       gs2 = gs_list$gs_negative)

# Merge by cluster
cL_resutls <- do.call("rbind", lapply(unique(seu@meta.data$IST_unsup_clus), function(cl) {
    es.max.cl = sort(rowSums(es.max[ ,rownames(seu@meta.data[seu@meta.data$IST_unsup_clus==cl, ])]), 
                     decreasing = !0)
    head(data.frame(cluster = cl, type = names(es.max.cl), scores = es.max.cl, 
                    ncells = sum(seu@meta.data$IST_unsup_clus==cl)), 10)
}))

sctype_scores <- cL_resutls %>% group_by(cluster) %>% top_n(n = 1, wt = scores)

# Set low-confident (low ScType score) clusters to "unknown"
sctype_scores$type[as.numeric(as.character(sctype_scores$scores)) < sctype_scores$ncells/4] <- "Unknown"

## Self coded addition
# Maintain original cluster separation
sctype_scores$cl_type <- paste0(sctype_scores$cluster, "_", sctype_scores$type)

# Sort by cluster
sctype_scores <- sctype_scores %>% arrange(as.numeric(sctype_scores$cluster))
print(sctype_scores)
```

In this case the "Brain" database has confidently annotated all clusters but has merged a lot clusters that got the same label: clusters a, b, d, e and j as GABAergic neurons; c and h as Dopaminergic neurons; g and q as Glutamatergic neurons; f and i as Oligodendrocytes; and o and p as Cancer stem cells.  

## Add annotations to Seurat object

In order not to loose information, both the scType annotation and a "cluster+scType" annotation will be saved in the seurat object.  

```{r AddAnn}
# Convert to data.frame and select columns
annot_df <- sctype_scores %>% 
  dplyr::select(cluster, type, cl_type) %>%
  dplyr::mutate(cluster = as.character(cluster))

# Create named vectors for easy mapping
scType_ann <- setNames(annot_df$type, annot_df$cluster)
cl_scType_ann <- setNames(annot_df$cl_type, annot_df$cluster)

# Add new metadata columns to seurat object
seu$scType_ann <- factor(scType_ann[as.character(seu$IST_unsup_clus)],
                         levels = unique(annot_df$type))

seu$cl_scType_ann <- factor(cl_scType_ann[as.character(seu$IST_unsup_clus)],
                            levels = annot_df$cl_type)

# Set the cl_scType annotation as default
Idents(seu) <- seu$scType_ann
```

```{r VizAnn, fig.width = 20, fig.height = 8}
# Save old custom palette for original clusters
n_cl <- n
cols_cl <- cols

# Create custom color palettes
n <- length(unique(Idents(seu)))
cols <- gg_color_hue(n)

# Inspect UMAP scType annotation
p1 <- DimPlot(seu, reduction = "umap", label = FALSE, raster = FALSE, 
              repel = TRUE, cols = cols_cl, group.by = "cl_scType_ann")

p2 <- DimPlot(seu, reduction = "umap", label = FALSE, raster = FALSE, 
              repel = TRUE, cols = cols, group.by = "scType_ann")

# Arrange plots
p1 + p2
```

## Save the Seurat object

```{r SavingSeuObj}
# Save the annotated Seurat object
name <- paste0("seu_", folder, "_ISTUnsup.RDS")
saveRDS(seu, here("output","processed_data",folder,name))
```

# Final spatial visualizations

Now that cells have been annotated, visualizing them in their spatial context becomes more informative. For example, all cells can be visualized simultaneously in the sample or one cell type can be specifically highlighted:     

```{r SpatialViz, fig.width = 13, fig.height = 7, , dpi = 300}
## Code adapted from Seurat Spatial vignette
p1 <- ImageDimPlot(seu, fov = "globalFOV", axes = TRUE, cols = cols) +
  theme_classic()

p2 <- ImageDimPlot(seu, fov = "globalFOV", axes = TRUE, cols = cols[3],
                   cells = WhichCells(seu, idents = "Oligodendrocytes")) +
  ggtitle("Oligodendrocytes") + 
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

# Arrange plots
p1 + p2
```

Another interesting visualization would be to plot gene expression markers in its spatial context. For example, one typical maker for Oligodendrocytes is *Olig1*:

```{r MarkerViz1, fig.width = 7, fig.height = 5}
## Code adapted from Seurat Spatial vignette
## Points are removed from violin plots as CosMx datasets can contain over 1 million cells and may hide the violin plot

VlnPlot(seu, features = "Olig1", pt.size = 0, y.max = 20) + NoLegend()
```
As expected, it is highly expressed in Oligodendrocytes and Oligodendrocyte precursor cells, but it can also be observed in its spatial context:

```{r SpatialViz2, fig.width = 13, fig.height = 7, dpi = 300}
## Code adapted from CosMxLite vignette
p1 <- ImageFeaturePlot(seu, fov = "globalFOV", features = "Olig1") +
  theme_classic() +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

p2 <- ImageDimPlot(seu, fov = "globalFOV", axes = TRUE, cols = cols[3],
                   cells = WhichCells(seu, idents = "Oligodendrocytes")) +
  ggtitle("Oligodendrocytes") + 
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", hjust = 0.5),
        legend.position = "bottom")

# Arrange plots
p1 + p2
```

# Performance and Session Info

<details>
  <summary>**Performance Report**</summary>
```{r PerformanceReport, echo = FALSE}
# Report dataframe
report_data <- data.table(
  Chunk = names(all_times),
  Time_sec = round(unlist(all_times), 2),
  Memory_Mb = round(unlist(all_mem), 2)
)

# Add total row
total_time <- round(sum(unlist(all_times)), 2)
total_mem <- round(sum(unlist(all_mem)), 2)

new_row <- list("Total", total_time, total_mem)
report_data <- rbindlist(list(report_data, new_row))

# Save as CSV
if(!dir.exists(here("output","performance_reports"))){
  dir.create(here("output","performance_reports"))
}

write.csv(report_data, here("output","performance_reports","4.1_insitutype_unsup_clustering_PR.csv"), row.names = FALSE)

# Show table
knitr::kable(report_data, row.names = FALSE, format = "html") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```
</details>


